Data exploration through dot-driven development\footnote{This work was supported by The Alan Turing Institute under the EPSRC grant EP/N510129/1 and by the Google Digital News Initiative.}

### 2017-05-22 ecoop-2017/paper.tex
Proceedings tweaks, mostly negative hspace, lolz

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{drop}(F) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\qquad\ball{1}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f \in \dom(F)~\textnormal{where}~C_f, L_f = \ident{drop}(F')\\
\quad \qquad \kvd{member}~\qident{drop $f$} : C_f = C_f(\Pi_{\ident{dom}(F')}(x)) & \quad\textnormal{and}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)\qquad\ball{2}                 &\textnormal{where}~C', L' = \ident{pivot}(F)\\
\\[0.5em]
\ident{sort}(F) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f) \}\qquad\ball{3}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\{f~|~F(f)=\kvd{num} \},~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{f \mapsto \ident{desc}})\\
\quad \qquad \kvd{member}~\qident{by $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{f \mapsto \ident{asc}})\\
\\[1em]
\multicolumn{2}{l}{ \ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup L') \}\qquad\ball{4} }\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-1em}\forall f\in\{f~|~F(f)=\kvd{num}, \,\nexists i.s_i=f'\mapsto\omega\wedge f'=f\}\\
\quad \qquad \kvd{member}~\qident{and $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{desc}})\\
\quad \qquad \kvd{member}~\qident{and $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{asc}})\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(\tau_{s_1,\ldots,s_n}(x))\quad~~\ball{5}           &\textnormal{where}~C', L' = \ident{pivot}(F)\\
\end{array}
\end{equation*}

@@

Types for defining the sorting transformation are split between two functions; \ident{sort}~\ball{3} 
generates type for choosing the first sorting key and \ident{sort-and}~\ball{4} lets the user add more
keys. The members are restricted to numerical columns (by checking $F(f)=\kvd{num}$).
The sort keys are kept as a vector. The \ident{sort} operation creates a singleton vector;
\ident{sort-and} appends a new key to the end and the \ident{then} member \ball{5} generates code 
that passes the collected sort keys to the $\tau$ operation of the relational algebra.
When generating members for adding further sort keys, we exclude the columns that are used
already (by checking that the column $f$ does not match column name of any of the existing 
keys $\nexists i.\,s_i = f'\mapsto\omega$).

@@

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{group}(F) = C, \{ C\mapsto (l, \bigcup L_f) \}\quad\ball{1} \\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
\\[0.5em]
\multicolumn{2}{l}{ \ident{agg}(F,f,G,S) = C, \{C\mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup\bigcup L''_f\cup L'\cup L'') \}\quad\ball{3}}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)\\
\quad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) & \quad \textnormal{when}~F(f)=\kvd{num}\quad\ball{4} \\
\quad \qquad \kvd{member}~\qident{concat $f$} : C''_f = C''_f(x) & \quad \textnormal{when}~F(f)=\kvd{string}\quad\ball{5} \\
\quad \qquad \kvd{member}~\qident{count all} : C' = C'(x) & \quad \textnormal{when}~\ident{Count}\notin G\quad\ball{6} \\
\quad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x)  \\
\quad \qquad \kvd{member}~\ident{then} : C'' = C''(\Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(x)) &\textnormal{where}~\{\rho_1/f_1,\ldots,\rho_n/f_n\}=S\quad\ball{7} \\[0.5em]
\qquad  \textnormal{where}\\[0.5em]
\multicolumn{2}{l}{
~\,\qquad \begin{array}{l}
 C_f, L_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num}\}, S\cup\{\ident{dist}~f/f\} )\\
 C'_f, L'_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num} \}, S\cup\{\ident{sum}~f/f\})\\
 C''_f, L''_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{string} \}, S\cup\{\ident{conc}~f/f\})\\
 C', L' = \ident{agg}(F,f, G\cup\{ \ident{Count}\mapsto\kvd{int} \}, S\cup{\ident{count}/\ident{Count}})\\
 C'', L'' = \ident{pivot}(G)
\end{array} 
}
\end{array}
\end{equation*}

@@

\begin{definition}[Expansion]
\label{thm:pivot-well}
Given a context $L$, we define $n^{\textnormal{th}}$ expansion of $L$, written 
$\downarrow_n L$ such that
$\downarrow_{n+1} L = \downarrow_{n} L \cup \bigcup L_n$ where $\downarrow_{n} L = \{ C_0\mapsto(l_0, L_0),\ldots,C_n\mapsto(l_n, L_n) \}$
and $\downarrow_0 L = L$.
\end{definition}

@@

\subsection{Related work}
\label{sec:further}

>>>>

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{drop}(F) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\qquad\ball{1}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f \in \dom(F)~\textnormal{where}~C_f, L_f = \ident{drop}(F')\\
\quad \qquad \kvd{member}~\qident{drop $f$} : C_f = C_f(\Pi_{\ident{dom}(F')}(x)) & \quad\textnormal{and}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)\qquad\ball{2}                 &\textnormal{where}~C', L' = \ident{pivot}(F)\\
\\[0.5em]
\ident{sort}(F) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f) \}\qquad\ball{3}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\{f~|~F(f)=\kvd{num} \},~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$ desc} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{f \mapsto \ident{desc}})\\
\quad \qquad \kvd{member}~\qident{by $f$ asc} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{f \mapsto \ident{asc}})\\
\\[1em]
\multicolumn{2}{l}{ \ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup L') \}\qquad\ball{4} }\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-4em}\forall f\in\{f~|~F(f)=\kvd{num}, \,\nexists i.s_i=f'\mapsto\omega\wedge f'=f\}\\
\quad \qquad \kvd{member}~\qident{$f$ desc} : C_f = C_f(x) &\hspace{-2em} C_f, L_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{desc}})\\
\quad \qquad \kvd{member}~\qident{$f$ asc} : C'_f = C'_f(x) &\hspace{-2em} C'_f, L'_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{asc}})\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(\tau_{s_1,\ldots,s_n}(x))  &\hspace{-2em}\textnormal{where}~C', L' = \ident{pivot}(F)\qquad ~~\ball{5}\\
\end{array}
\end{equation*}

@@

Types for defining the sorting transformation are split between two functions; \ident{sort}~\ball{3} 
generates type for choosing the first sorting key and \ident{sort-and}~\ball{4} lets the user add more
keys. For space reasons, we abbreviate \ident{ascending} and \ident{descending} as \ident{asc} and
\ident{desc} in the generated member names and we omit \ident{and} in name of further keys such as 
\qident{and Gold descending}.

The members are restricted to numerical columns (by checking $F(f)=\kvd{num}$).
The sort keys are kept as a vector. The \ident{sort} operation creates a singleton vector;
\ident{sort-and} appends a new key to the end and the \ident{then} member \ball{5} generates code 
that passes the collected sort keys to the $\tau$ operation of the relational algebra.
When generating members for adding further sort keys, we exclude the columns that are used
already (by checking that the column $f$ does not match column name of any of the existing 
keys $\nexists i.\,s_i = f'\mapsto\omega$).

@@

\begin{figure}
\begin{equation*}
\hspace{-1.5em}
\begin{array}{ll}
\ident{group}(F) = C, \{ C\mapsto (l, \bigcup L_f) \}\quad\ball{1} \\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-2.5em}\forall f\in\dom(F)~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\hspace{-2.5em}\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
\\[0.5em]
\multicolumn{2}{l}{ \ident{agg}(F,f,G,S) = C, \{C\mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup\bigcup L''_f\cup L'\cup L'') \}\quad\ball{3}}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)\\
\quad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) & \quad \textnormal{when}~F(f)=\kvd{num}\quad\ball{4} \\
\quad \qquad \kvd{member}~\qident{concat $f$} : C''_f = C''_f(x) & \quad \textnormal{when}~F(f)=\kvd{string}\quad\ball{5} \\
\quad \qquad \kvd{member}~\qident{count all} : C' = C'(x) & \quad \textnormal{when}~\ident{Count}\notin G\quad\ball{6} \\
\quad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x)  \\
\quad \qquad \kvd{member}~\ident{then} : C'' = C''(\Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(x)) &\textnormal{where}~\{\rho_1/f_1,\ldots,\rho_n/f_n\}=S\quad\ball{7} \\[0.5em]
\qquad  \textnormal{where}\\[0.5em]
\multicolumn{2}{l}{
~\,\qquad \begin{array}{l}
 C_f, L_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num}\}, S\cup\{\ident{dist}~f/f\} )\\
 C'_f, L'_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num} \}, S\cup\{\ident{sum}~f/f\})\\
 C''_f, L''_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{string} \}, S\cup\{\ident{conc}~f/f\})\\
 C', L' = \ident{agg}(F,f, G\cup\{ \ident{Count}\mapsto\kvd{int} \}, S\cup{\ident{count}/\ident{Count}})\\
 C'', L'' = \ident{pivot}(G)
\end{array} 
}
\end{array}
\end{equation*}

@@

\begin{definition}[Expansion]
\label{thm:pivot-expand}
Given a context $L$, we define $n^{\textnormal{th}}$ expansion of $L$, written 
$\downarrow_n L$ such that
$\downarrow_{n+1} L = \downarrow_{n} L \cup \bigcup L_n$ where $\downarrow_{n} L = \{ C_0\mapsto(l_0, L_0),\ldots,C_n\mapsto(l_n, L_n) \}$
and $\downarrow_0 L = L$.
\end{definition}

@@

\subsection{Related work}
\label{sec:related}
